Forest canopies are better generators of biodiversity in the tropics

### 2025-06-27 manuscript.Rmd
revisions

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. As vertical, topographically complex landscape features, mountains provide a larger volume of geographic space and produce steep climate gradients over short distance that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013]. Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

Tall and structurally complex forests have been proposed to generate biodiversity through a similar mechanism to mountains. Tall, multi-layered forests increase habitat volume and contain vertical microclimate and resource profiles, which allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, as forest height increases and provides more habitable space, additional species may join the assemblage, leading to greater richness within a given area compared to shorter forests (cite). This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex (cite). Yet relationships between canopy height or volume and species richness are often weak in temperate and boreal forests, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015] - a pattern resembling Humbolt's Enigma in forests.

Climate may play a key role in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles. Vertical microclimate gradients produced by micrometeorological processes involving interactions between macroclimate, solar radiation, wind patterns, and forest structure are characterized by decreasing humidity and increasing thermal variability from the ground to the canopy and may help or hinder the evolution of diverse arboreal assemblages. In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in tropical regions could thus support higher animal verticality and richness. In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite). Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.

@@

>>>>

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains as vertical and topographically complex landscape features in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. The vertical nature of mountains provides a larger volume of geographic space and produces steep climate gradients over short distances that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013]. Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

Tall, multi-layered forests have been proposed to generate biodiversity through a similar mechanism to mountains. By increasing habitat volume and producing vertical profiles of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to shorter forests [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Yet impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020] - a pattern resembling that of mountains.

Climate and vegetative complexity may play key roles in mediating the relationships between canopy height and species richness by physiologically constraining arboreal lifestyles and impacting the availability of vertical niche opportunities. Vertical microclimate gradients produced by micrometeorological processes involving interactions between macroclimate, solar radiation, wind patterns, and forest structure are characterized by decreasing humidity and increasing thermal variability from the ground to the canopy and may help or hinder the evolution of diverse arboreal assemblages. In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to finely partition the entire three-dimensional habitat space [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in tropical regions could thus support higher animal verticality and richness. In contrast, extreme cold, or the combination of heat and drought may impede the evolution of arboreal lifestyles, constraining biodiversity to lower vertical positions tat are more climatically buffered (cite). Harsh macroclimate conditions may therefore truncate the availability of vertical habitat space, thereby limiting vertical stratification and species richness despite the presence of vegetation.

@@

# Methods
## Data
The canopy height layer was obtained from Lang et al. [-@langHighresolutionCanopyHeight2023], which represents canopy height in the year 2020 at a 10 m resolution. To obtain canopy height in forests, we aggregated the data to match the resolution of Hansen et al. [-@hansenGlobalLandUse2022] global land cover and land use in 2019 (~ 30 m resolution) and used this layer to mask all grid cells with woody vegetation < 3 m, surface water and permanent ice, built-up land, cropland, recent forest loss, and land use associated with forestry and shifting cultivation (e.g., oil palm). We then aggregated and averaged the remaining pixels to the resolution of the verticality data (50 km).
